---
output: github_document
---


# virtuoso <img src="man/figures/logo.svg" align="right" alt="" width="120" />

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/cboettig/virtuoso.svg?branch=master)](https://travis-ci.org/cboettig/virtuoso)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/cboettig/virtuoso?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/virtuoso)
[![Coverage status](https://codecov.io/gh/cboettig/virtuoso/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/virtuoso?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/virtuoso)](https://cran.r-project.org/package=virtuoso) [![](http://badges.ropensci.org/271_status.svg)](https://github.com/ropensci/sofware-review/issues/271)

 
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


The goal of virtuoso is to provide an easy interface to Virtuoso RDF database from R. 

## Installation

You can install the development version of virtuoso from GitHub with:

``` r
remotes::install_github("cboettig/virtuoso")
```

## Getting Started

```{r}
library(virtuoso)
```

For Mac users, `virtuoso` package includes a utility function to install and configure a local Virtuoso Open Source instance using Homebrew.  Otherwise, simply install the Virtuoso Open Source edition for your operating system. 


```{r install}
vos_install()
```

We can now start our Virtuoso server from R:

```{r}
vos_start()
```



 Once the server is running, we can connect to the database.

```{r}
con <- vos_connect()
```

Our connection is now live, and accepts SPARQL queries directly.  

```{r}
DBI::dbGetQuery(con, "SPARQL SELECT * WHERE { ?s ?p ?o } LIMIT 4")
```

## DSL

`virtuoso` also provides wrappers around some common queries to make it easier to work with Virtuoso and RDF.

The bulk loader can be used to quickly import existing sets of triples. 

```{r}
example <- system.file("extdata", "person.nq", package = "virtuoso")
vos_import(con, example)
```

Can also read in compressed formats as well.  Remember to set the pattern match appropriately.  This is convenient because N-Quads compress particularly well, often by a factor of 20 (or rather, can be particularly large when uncompressed, owing to the repeated property and subject URIs).  

```{r}
ex <- system.file("extdata", "library.nq.gz", package = "virtuoso")
vos_import(con, ex)
```

`vos_import` invisibly returns a table of the loaded files, with error message and loading times.  If a file cannot be imported, an error message is returned:

```{r error = TRUE}
bad_file <- system.file("extdata", "bad_quads.nq", package = "virtuoso")
vos_import(con, bad_file)
```

 
We can now query the imported data using SPARQL.  

```{r}
vos_query(con, 
"SELECT ?p ?o 
 WHERE { ?s ?p ?o .
        ?s a <http://schema.org/Person>
       }")
```

```{r}
vos_query(con, 
"SELECT ?p ?o 
 WHERE { ?s ?p ?o .
        ?s a <http://example.org/vocab#Chapter>
       }")
```


## Server controls

We can control any `virtuoso` server started with `vos_start()` using a series of helper commands.  

```{r}
vos_status()
```

Advanced usage note: `vos_start()` invisibly returns a `processx` object which we can pass to other server control functions, or access the embedded `processx` control methods directly.  The `virtuoso` package also caches this object in an environment so that it can be accessed directly without having to keep track of an object in the global environment. Use `vos_process()` to return the `processx` object.  For example:

```{r}
library(ps)
p <- vos_process()
ps_is_running(p)
ps_cpu_times(p)
ps_suspend(p)
ps_resume(p)
```

```{r include = FALSE}
vos_kill()
```

## Going further

Please see the package vignettes for more information:

- [details on Virtuoso Installation & configuration](https://cboettig.github.io/virtuoso/articles/installation.html)
- [The Data Lake: richer examples of RDF use](https://cboettig.github.io/virtuoso/articles/articles/datalake.html)

---

Please note that the `virtuoso` R package is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.


```{r include=FALSE}
if(require(codemetar)) codemetar::write_codemeta()
```
