---
output: github_document
---

 [![Travis build status](https://travis-ci.org/cboettig/virtuoso.svg?branch=master)](https://travis-ci.org/cboettig/virtuoso)
 
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# virtuoso

The goal of virtuoso is to provide an easy interface to Virtuoso RDF database from R. 

## Installation

You can install the development version of virtuoso from GitHub with:

``` r
remotes::install_github("cboettig/virtuoso")
```

## Getting Started

```{r}
library(virtuoso)
```

For Mac users, `virtuoso` package includes a utility function to install and configure a local Virtuoso Open Source instance using Homebrew.  Otherwise, simply install the Virtuoso Open Source edition for your operating system. 


```{r install}
vos_install()
```

We can now start a server.  (Here we assign the server process to an object, `myserver` which we can use to stop or control it explicitly, if necessary). Note that the server may take a few seconds to come up.  

```{r}
myserver <- vos_start()

Sys.sleep(5)
```



 Once the server is running, we can connect to the database.

```{r}
con <- vos_connect()
```

Our connection is now live, and accepts SPARQL queries directly.  

```{r}
ex <- DBI::dbGetQuery(con, "SPARQL SELECT * WHERE { ?s ?p ?o }")
```

## DSL

`virtuoso` also provides wrappers around some common queries to make it easier to work with Virtuoso and RDF.

The bulk loader can be used to quickly import existing sets of triples. 

```{r}
example <- system.file("extdata", "person.nq", package = "virtuoso")
vos_import(con, example)
```

We can clear all data in the default graph if we want a fresh start:

```{r}
vos_clear_graph(con)
```

---

See richer examples in the package vignettes.


