---
output: github_document
---

 [![Travis build status](https://travis-ci.org/cboettig/virtuoso.svg?branch=master)](https://travis-ci.org/cboettig/virtuoso)
 
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = TRUE
)
```
# virtuoso

The goal of virtuoso is to provide an easy interface to Virtuoso RDF database from R. 

## Installation

You can install the development version of virtuoso from GitHub with:

``` r
remotes::install_github("cboettig/virtuoso")
```

## Getting Started

```{r}
library(virtuoso)
```

For Mac users, `virtuoso` package includes a utility function to install and configure a local Virtuoso Open Source instance using Homebrew.  Otherwise, simply install the Virtuoso Open Source edition for your operating system. 


```{r install}
vos_install()
```

We can now start a server.  (Here we assign the server process to an object, `myserver` which we can use to stop or control it explicitly, if necessary). Note that the server may take a few seconds to come up.  

```{r}
myserver <- vos_start()

Sys.sleep(5)
```



 Once the server is running, we can connect to the database.

```{r}
con <- vos_connect()
```

Our connection is now live, and accepts SPARQL queries directly.  

```{r}
ex <- DBI::dbGetQuery(con, "SPARQL SELECT * WHERE { ?s ?p ?o }")
head(ex)
```

## DSL

`virtuoso` also provides wrappers around some common queries to make it easier to work with Virtuoso and RDF.

The bulk loader can be used to quickly import existing sets of triples. 

```{r}
example <- system.file("extdata", "person.nq", package = "virtuoso")
vos_import(con, example)
```

We can clear all data in the default graph if we want a fresh start:

```{r}
vos_clear_graph(con)
```

-----


## Tabular data as RDF?


We can represent any data as RDF with a little care.  For instance, consider the `nycflights13` data. 

```{r}
library(nycflights13)
library(dplyr)

```


First, we must represent any primary or foreign keys in any table must be as URIs and not literal integers or strings: 

```{r as_uri}
uri_flights <- flights %>% 
  mutate(tailnum = paste0("planes:", tailnum),
         carrier = paste0("airlines:", carrier))
```

We write the `data.frame`s out as nquads.  Recall that each cell of a `data.frame` can be represented as a triple, in which the column is the predicate, the primary key (or row number) the subject, and the cell value the object.  We turn column names and primary keys into URIs using a prefix based on the table name.  (Note that `rdflib` does this conversion by merely munging cells and calling `write.table`, it is not a standard `redland` library transform).

```{r}
write_nquads(airlines,  "airlines.nq", key = "carrier", prefix = "airlines:")
write_nquads(planes,  "planes.nq", key = "tailnum", prefix = "planes:")
write_nquads(uri_flights,  "flights.nq", prefix = "flights:")
```

We're ready to import all these triples.  This may take a few minutes:

```{r}
system.time(
  vos_import(con, c("flights.nq", "planes.nq", "airlines.nq"))
)
```


The data from all three tables is now reduced into a single triplestore graph, one triple for each data point. Rather than joining tables, we can write SPARQL query that names the columns we want.



```{r}

query <- 
'SELECT  ?carrier ?name ?manufacturer ?model ?dep_delay
WHERE {
?flight <flights:tailnum>  ?tailnum .
?flight <flights:carrier>  ?carrier .
?flight <flights:dep_delay>  ?dep_delay .
?tailnum <planes:manufacturer> ?manufacturer .
?tailnum <planes:model> ?model .
?carrier <airlines:name> ?name
}'

system.time(
df <- vos_query(con, query)
)

head(df)
```


