---
output: github_document
---

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/cboettig/virtuoso.svg?branch=master)](https://travis-ci.org/cboettig/virtuoso)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/cboettig/virtuoso?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/virtuoso)
[![Coverage status](https://codecov.io/gh/cboettig/virtuoso/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/virtuoso?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/virtuoso)](https://cran.r-project.org/package=virtuoso) 
 
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# virtuoso

The goal of virtuoso is to provide an easy interface to Virtuoso RDF database from R. 

## Installation

You can install the development version of virtuoso from GitHub with:

``` r
remotes::install_github("cboettig/virtuoso")
```

## Getting Started

```{r}
library(virtuoso)
```

For Mac users, `virtuoso` package includes a utility function to install and configure a local Virtuoso Open Source instance using Homebrew.  Otherwise, simply install the Virtuoso Open Source edition for your operating system. 


```{r install}
vos_install()
```

We can now start our Virtuoso server from R:

```{r}
vos_start()
Sys.sleep(10)

```



 Once the server is running, we can connect to the database.

```{r}
con <- vos_connect()
```

Our connection is now live, and accepts SPARQL queries directly.  

```{r}
DBI::dbGetQuery(con, "SPARQL SELECT * WHERE { ?s ?p ?o } LIMIT 4")
```

## DSL

`virtuoso` also provides wrappers around some common queries to make it easier to work with Virtuoso and RDF.

The bulk loader can be used to quickly import existing sets of triples. 

```{r}
example <- system.file("extdata", "person.nq", package = "virtuoso")
vos_import(con, example)
```

Can also read in compressed formats as well.  Remeber to set the pattern match appropriately.  This is convient becuase N-Quads compress particularly well, often by a factor of 20 (or rather, can be particularly large when uncompressed, owing to the repeated property and subject URIs).  

```{r}
ex <- system.file("extdata", "library.nq.gz", package = "virtuoso")
vos_import(con, ex, ext = "*.nq.gz")
```

The import process is run by the external process, it will not throw an error if the server fails to import the file (e.g. due to formatting errors in the N-Quads file).  We can optionally check for possible error messages in the import process by scanning the full log for errors:

```{r}
vos_log(just_errors = TRUE)
```

We can now query the imported data using SPARQL.  

```{r}
vos_query(con, 
"SELECT ?p ?o 
 WHERE { ?s ?p ?o .
        ?s a <http://schema.org/Person>
       }")
```

```{r}
vos_query(con, 
"SELECT ?p ?o 
 WHERE { ?s ?p ?o .
        ?s a <http://example.org/vocab#Chapter>
       }")
```


## Server controls

We can control any `virtuoso` server started with `vos_start()` using a series of helper commands.  

```{r}
vos_status()
```

Advanced usage note: `vos_start()` invisibly returns a `processx` object which we can pass to other server control functions, or access the embedded `processx` control methods directly.  The `virtuoso` package also caches this object in an environment so that it can be accessed directly without having to keep track of an object in the global environment. Use `vos_process()` to return the `processx` object.  For example:

```{r}
p <- vos_process()
p$get_error_file()
p$suspend()
p$resume()
```

```{r include = FALSE}
vos_kill()
```

---

See richer examples in the [package vignettes](https://cboettig.github.io/virtuoso/articles).


