---
title: "A Brief Introduction to Virtuoso"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{readme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# virtuoso

The goal of virtuoso is to provide an easy interface to Virtuoso RDF database from R. 

## Installation

You can install the development version of `virtuoso` from GitHub with:

``` r
remotes::install_github("cboettig/virtuoso")
```

Linux users should install Virtuoso server from their package manager.


## Getting Started

```{r}
library(virtuoso)
```

For Mac users, `virtuoso` package includes a utility function to install and configure a local Virtuoso Open Source instance using Homebrew.  Otherwise, simply install the Virtuoso Open Source edition for your operating system. 


```{r install}
vos_install()
```

We can now start our Virtuoso server from R:

```{r}
vos_start()
```



 Once the server is running, we can connect to the database.

```{r}
con <- vos_connect()
```

Our connection is now live, and accepts SPARQL queries directly.  

```{r}
ex <- DBI::dbGetQuery(con, "SPARQL SELECT * WHERE { ?s ?p ?o }")
```

## DSL

`virtuoso` also provides wrappers around some common queries to make it easier to work with Virtuoso and RDF.

The bulk loader can be used to quickly import existing sets of triples. 

```{r}
example <- system.file("extdata", "person.nq", package = "virtuoso")
vos_import(con, example)
```

```{r}
vos_query(con, 
"SELECT ?p ?o 
 WHERE { ?s ?p ?o .
        ?s a <http://schema.org/Person>
       }")
```

We can clear all data in the default graph if we want a fresh start:

```{r}
vos_clear_graph(con)
```



## Server controls

We can control any `virtuoso` server started with `vos_start()` using a series of helper commands.  

```{r}
vos_status()
```

Advanced usage note: `vos_start()` invisibly returns a `processx` object which we can pass to other server control functions, or access the embedded `processx` control methods directly.  The `virtuoso` package also caches this object in an environment so that it can be accessed directly without having to keep track of an object in the global environment. Use `vos_process()` to return the `processx` object.  For example:

```{r}
p <- vos_process()
p$get_error_file()
p$suspend()
p$resume()
```

```{r include = FALSE}
vos_kill()
```

---

See richer examples in the package vignettes.


