---
title: "The Data Lake: Schema on Read"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Datal Lake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



```{r}
library(virtuoso)
library(rdftools) # for write_nquads

library(nycflights13)
library(dplyr)

```

We start up our Virtuoso server, wait for it to come up, and then connect:

```{r install}
myserver <- vos_start()
Sys.sleep(5)
con <- vos_connect()
```


We can represent any data as RDF with a little care.  For instance, consider the `nycflights13` data. First, we must represent any primary or foreign keys in any table must be as URIs and not literal integers or strings: 

```{r as_uri}
uri_flights <- flights %>% 
  mutate(tailnum = paste0("planes:", tailnum),
         carrier = paste0("airlines:", carrier))
```

We write the `data.frame`s out as nquads.  Recall that each cell of a `data.frame` can be represented as a triple, in which the column is the predicate, the primary key (or row number) the subject, and the cell value the object.  We turn column names and primary keys into URIs using a prefix based on the table name.  (Note that `rdflib` does this conversion by merely munging cells and calling `write.table`, it is not a standard `redland` library transform).

```{r}
write_nquads(airlines,  "airlines.nq", key = "carrier", prefix = "airlines:")
write_nquads(planes,  "planes.nq", key = "tailnum", prefix = "planes:")
write_nquads(uri_flights,  "flights.nq", prefix = "flights:")
```

We're ready to import all these triples.  This may take a few minutes:

```{r}
system.time(
  vos_import(con, c("flights.nq", "planes.nq", "airlines.nq"))
)
```


The data from all three tables is now reduced into a single triplestore graph, one triple for each data point. Rather than joining tables, we can write SPARQL query that names the columns we want.



```{r}
query <- 
'SELECT  ?carrier ?name ?manufacturer ?model ?dep_delay
WHERE {
?flight <flights:tailnum>  ?tailnum .
?flight <flights:carrier>  ?carrier .
?flight <flights:dep_delay>  ?dep_delay .
?tailnum <planes:manufacturer> ?manufacturer .
?tailnum <planes:model> ?model .
?carrier <airlines:name> ?name
}'

system.time(
df <- vos_query(con, query)
)

head(df)
```




```{r}
myserver$kill()
```
